---
title: "Habitat complexity part 1. Group clustering"
author: "Alberto Rovellini"
date: "18 March 2020"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

This document produces a distance matrix based on measures of structural complexity, and then applies cluster analysis. This document then parts the quadrats into high and low complexity based on the structural complexity metrics.

```{r, echo = FALSE, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, width = 150)
```

```{r, message = FALSE, warning = FALSE}

library(readxl)
library(data.table)
library(ggplot2)
library(abind)
library(dplyr)
library(reshape2)
library(RColorBrewer)
library(emmeans)
library(ggfortify)
library(vegan)
library(BiodiversityR)
# library(ggvegan)
library(cluster)
library(factoextra)
library(magrittr)

```

Read in cover data.

```{r}

cover <- read.csv(".../cover.csv")

```

Merge in results of the 3D models. This contains the model-derived structural metrics (*R*, 1/*k*, *D*~1, *D*~2, *D*~3, *D*~4, *D*~5).

```{r}

hcx <- read.csv(".../hcx_start.csv")

hcx.add <- hcx %>% dplyr::select(Rugosity.AV:lnS6)
n.times <- length(levels(droplevels(cover$Long_group))) # find number of times one row og hcx must be replicated to fit with cover

hcx.add <- hcx.add[rep(seq_len(nrow(hcx.add)), each = n.times),] # inflate hcx

cover.hcx <- cbind(cover, hcx.add) # stitch

```

Here we build a distance matrix on the structural properties. We standardize the initial multivariate matrix.

## Model-derived metrics.

```{r, fig.width = 6, fig.height = 5}

hcx.matrix <- hcx %>% dplyr::select(
  Rugosity.AV, K.AV, F1:F5
) %>% mutate_all(
  funs(scale)
)

scale(hcx$Rugosity.AV)

rownames(hcx.matrix) <- hcx$Quadrat

distmat <- vegdist(hcx.matrix, method = "euclidean") # eculidean space

fviz_dist(distmat, 
   gradient = list(low = "firebrick", mid = "white", high = "dodgerblue")) # this visualises the whole distance matrix

```

### Cluster analysis

This is based on the matrix of the euclidean distances derived from the the multivariate matrix *X* of model-derived structural complexity metrics *R*, 1/*k*, *D*~1-5~. The multivariate matrix is first standardised (for each x~ij~ value of *X*, (x~ij~-colmean)/colsd), so that all columns are on the same scale. Then agglomerative hierarchical cluster analysis with Ward's method (Ward 1963, Murtagh and Legendre 2014) is applied. 

## Corals

```{r, fig.width = 5, fig.height = 4.5}

# prepare distance matrix Y

hcx.matrix.coral <- hcx.matrix[-grep("S", rownames(hcx.matrix)),]

# prepare data matrix X

cover.matrix <- read.csv(".../coverShortFormat.csv")

rownames(cover.matrix) <- cover.matrix$X

cover.coral <- cover.matrix[-which(grepl("S", rownames(cover.matrix))), -1]

distmat <- dist(hcx.matrix.coral, method = "euclidean") 

coral.clusts <- hclust(distmat, method = "ward.D2") # hierarchical clustering method

coral.dendro <- fviz_dend(coral.clusts, k = 2, k_colors = c("firebrick", "dodgerblue"), rect = TRUE, lwd = 1, horiz = T)

coral.dendro

```

## Sponges

```{r, fig.width = 5, fig.height = 4.5}

# prepare distance matrix Y

hcx.matrix.sponge <- hcx.matrix[grep("S", rownames(hcx.matrix)),]

# prepare data matrix X

cover.sponge <- cover.matrix[which(grepl("S", rownames(cover.matrix))), -1]

distmat <- dist(hcx.matrix.sponge, method = "euclidean") 

sponge.clusts <- hclust(distmat, method = "ward.D2") # hierarchical clustering method

sponge.dendro <- fviz_dend(sponge.clusts, k = 2, k_colors = c("firebrick", "dodgerblue"), rect = TRUE, lwd = 1, horiz = T)

sponge.dendro

```

# Data regrouping

Add the new classification to the quadrats names. 

```{r}

cluster.order.coral <- coral.clusts$labels[coral.clusts$order] # this extracts the order from low to high complexity
cluster.order.sponge <- sponge.clusts$labels[sponge.clusts$order]

# make vector for coral samples (manually set to 9 as cutoff low-high, as corals are 9 L and 6 H)

cluster.order.coral[1:9] <- paste(substr(cluster.order.coral[1:9], 1, nchar(cluster.order.coral[1:9])-2), "CL", sep = "_")
cluster.order.coral[10:length(cluster.order.coral)] <- paste(substr(cluster.order.coral[10:length(cluster.order.coral)], 1, nchar(cluster.order.coral[9:length(cluster.order.coral)])-2), "CH", sep = "_")

# make vector for sponge samples (manually set to 9 as cutoff low-high, as sponges are 9 L and 8 H)

cluster.order.sponge[1:9] <- paste(substr(cluster.order.sponge[1:9], 1, nchar(cluster.order.sponge[1:9])-2), "SL", sep = "_")
cluster.order.sponge[10:length(cluster.order.sponge)] <- paste(substr(cluster.order.sponge[10:length(cluster.order.sponge)], 1, nchar(cluster.order.sponge[9:length(cluster.order.sponge)])-2), "SH", sep = "_")

# concatenate and order them as the original

new.groups <- c(cluster.order.coral, cluster.order.sponge)

new.groups <- sort(new.groups)

# for quadrats 3 and 15 2 "_" are concatenated, unclear why. Manually fix

new.groups <- gsub("__", "_", new.groups)

# also make label vector for plots

labels <- paste(substr(new.groups, nchar(new.groups)-1, nchar(new.groups)), "_", substr(new.groups, 4, 6), sep = "")

```

Add this vector to: cover and hcx. Write out a csv file each.

```{r}

# hcx

hcx <- hcx %>% dplyr::mutate(
  Quadrat    = new.groups,
  Label      = labels,
  Complexity = substr(new.groups, nchar(new.groups), nchar(new.groups))
)

write.csv(hcx, ".../hcxNew.csv", row.names = FALSE)

# cover

cover <- cover %>% dplyr::mutate(
  Quadrat    = rep(new.groups, each = length(levels(factor(cover$Long_group)))),
  Label      = rep(labels, each = length(levels(factor(cover$Long_group)))),
  Complexity = substr(Quadrat, nchar(Quadrat), nchar(Quadrat))
) 

write.csv(cover, ".../coverNew.csv", row.names = FALSE)

```
